Thermal reactionomes reveal divergent responses to thermal extremes in warm

### 2015-11-05 ApTranscriptome_TR.Rmd
revisions

---
title: Technical Report for **Thermal reactionomes reveal divergent responses to thermal
  extremes in warm and cool-climate ant species**
author: '[John Stanton-Geddes](john.stantongeddes.research@gmail.com)'
output:
  html_document:
    keep_md: yes
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  word_document: default
---

@@

# table
(Ap.type.table <- rbind(A22.type.table, Ar.type.table))

@@

```{r rData, echo=FALSE}
# save RData for manuscript file
save(Ap.type.table, annotation.table, annotation.table.responsive, Ap.type.table.prop, 
     chi1, chi2, chi3, chi4, chi5, chi6, D, mh.test, pre.signif.transcripts,  
     responsive_transcripts_w_BLAST, sigtable, signif.transcripts, T_on.low.ttest, 
     T_on.high.ttest, T_off.int.high.ttest, T_off.int.low.ttest, w3, 
     file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))
```

>>>>

---
title: Technical Report for **Thermal reactionomes reveal divergent responses to thermal
  extremes in warm and cool-climate ant species**
author: '[John Stanton-Geddes](john.stantongeddes.research@gmail.com)'
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document:
    keep_md: yes
---

@@

# table
(Ap.type.table <- rbind(A22.type.table, Ar.type.table))
rownames(Ap.type.table) <- list("A. picea", "A. carolinensis")

@@

```{r rData, echo=FALSE}
# save RData for manuscript file
save(Ap.type.table, annotation.table, annotation.table.responsive, Ap.type.table.prop, 
     chi1, chi2, chi3, chi4, chi5, chi6, D, mh.test, pre.signif.transcripts,  
     responsive_transcripts_w_BLAST, sigtable, signif.transcripts, T_on.low.ttest, 
     T_on.high.ttest, T_off.int.high.ttest, T_off.int.low.ttest, w3, tt2,
     file = paste("doc/ms_", Sys.Date(), ".RData", sep = ""))
```

### 2015-11-05 doc/ApTranscriptome_manuscript.Rmd
revisions

The distributions of species and their responses to climate change are in part determined by their thermal tolerances. However, little is known about how thermal tolerance evolves. To test whether evolutionary extension of thermal limits is accomplished through enhanced cellular stress response (*enhanced response*), constitutively elevated expression of protective genes (*genetic assimilation*) or a shift from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*: the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized thermal reactionomes of two common ant species in the eastern U.S, the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*, across 12 temperatures that spanned their entire thermal breadth. We found that at least 2% of all genes were thermally responsive. The majority of upregulation was specific to exposure to low temperatures. The cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than did *A. carolinensis*, consistent with the *enhanced response* hypothesis. In contrast, under high temperatures the warm-adapted *A. carolinensis* downregulated many of the genes upregulated in *A. picea*, and required more extreme temperatures to induce down-regulation in gene expression, consistent with the *tolerance* hypothesis. We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*. These results suggest that increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.

@@

The combined *Aphaenogaster* transcriptome assembly contained `r prettyNum(nrow(annotation.table), big.mark =",")` transcripts. About half of these (`r prettyNum(nrow(annotation.table.responsive), big.mark =",")`) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. The resampling analysis revealed that the polynomial model was prone to overfitting the data. The 95th percentile of the resampled data returned a null expectation of `r quantile(D$num_signif, probs = 0.95)` putatively significant transcripts, when we would expect no significant relationship. With the true data, there were `r nrow(pre.signif.transcripts)` transcripts below the 5% FDR threshold, of which we retained only the top `r nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)` (`r round((nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)) / nrow(annotation.table), 3) * 100`)% as true positives given the null expectation from the resampling. Of these transcripts, `r round(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"] / sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), 2) *100`% (`r prettyNum(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"], big.mark=",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g. high vs. low temperature). The proportion of responsive transcripts is similar if we focus only on those transcript with a BLAST hit (`r prettyNum(responsive_transcripts_w_BLAST, big.mark = ",")` significant transcripts out of `r prettyNum(nrow(annotation.table.responsive), big.mark = ",")`). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] so this estimate is likely a lower bound for the true proportion of transcripts that are thermally responsive.

@@

In addition, we also examined the specific patterns of shifts from one expression category to another between species. When controlling for overall differences in the number of responsive transcripts, the expression categories of individual transcripts between the two species were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, P < 0.001, Fig. S1). Specifically, expression patterns, which in slope as well as category, differed significantly between species for `r prettyNum(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"], big.mark = ",")` (`r round(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"] / sigtable[which(sigtable$Coefficient=="Total"), "Number.significant"], 2) * 100`%) of the thermally responsive transcripts, most of which shifted from one expression category to another.

@@

Similarly, for down-regulated (**Intermediate**) transcripts, we compared the mean temperatures of downregulation of transcript expression between species at both high (> 20°C) and low (< 20°C) temperatures. Consistent with the *tolerance hypothesis*, *A. carolinensis* had greater inertia of gene expression in response to increasing temperatures. The temperature of downregulation for **Intermediate** transcripts was `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`). The difference between temperatures was not significant with decreasing temperatures (*t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig. 3B).

@@

We tested the *genetic assimilation hypothesis* by comparing the log ratios of relative inducibility to relative baseline expression at the rearing temperature (25°C). If stress-response transcripts have shifted between species from inducible to constitutive expression, there should be a negative relationship between the two. We found no evidence of such a relationship for either temperature extreme: transcripts more upregulated at high temperatures in the cool-climate *A. picea* were not expressed at higher baseline levels in the warm-climate *A. carolinensis* (Fig. 4A). Similarly, transcripts more upregulated at low temperatures in *A. carolinensis* did not show higher baseline levels in *A. picea* (Fig. 4B). In fact, for both comparisons we found a weakly positive relationship between relative inducibility and baseline expression between the two species ($\beta_1$ = 0.31, *P* < 0.001 and ($\beta_1$ = 0.21, *P* < 0.001). In addition, the thermally responsive transcripts in *A. carolinensis*, regardless of expression pattern, had higher baseline expression than those in *A. picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`). An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A. picea* but were not thermally responsive in *A. carolinensis* (top-row of Fig. 2B). These transcripts are less likely to be upregulated in any context, consistent with the tolerance hypothesis.

@@

To limit differences in gene expression not related to the experimental treatment, we haphazardly collected three ants from each 2012 colony at the same time of day on 12 different days. We measured response to temperature with a one-hour static temperature application, which is ecologically relevant for workers that leave the thermally-buffered nest and are immediately exposed to ambient temperatures while foraging [@lubertazzi2012]. Each day, the ants were placed in glass tubes immersed in a water bath maintained at one of 12 randomly-assigned temperatures (0° to 38.5°C, in 3.5° increments) for one hour. The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum (~0°C) and maximum (~43°C) temperatures for *Aphaenogaster* [@warren2013], and these treatments did not cause mortality. At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C. Thus, our reactionome characterized early, but not late, responding genes. We extracted mRNA by homogenizing the three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.

@@

We assembled the sequenced reads into the full set of mRNA transcripts, the transcriptome, for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011]. *De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g. regions of two separate transcripts that assemble into a false, or chimeric, third transcript) [@yang2013]. We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length. Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supporting Information). We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly. To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmieder2011] with the provided bacteria, virus, archaen and human databases of contaminants.

@@

We quantified expression of each transcript using the program `Sailfish` [@patro2014] and used the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of transcript expression. We included the contaminant transcripts identified by `DeconSeq` at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses. Because preliminary examination of the data (Supporting Information) indicated that the 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supporting Information).

@@

All analyses were performed with R 3.2 [@r_core_team2013] and are fully reproducible (Supporting Information).

@@

@@

The reproducible and version-controlled scripts for data analysis are available on GitHub (https://github.com/johnstantongeddes/ApTranscriptome).

## Author Contributions

@@

**Figure 1**. Prediction illustrations of expected thermal response patterns in the two species under alternative mechanistic hypotheses of temperature adaptation. Although both temperature extremes were investigated in a similar way, for simplicity only the response to low temperatures is illustrated here. Each column indicates the distribution across all response categories in *A. picea*, which has a lower CT~min~ and is therefore better adapated to low temperatures, for the set of transcripts identified as cold-induced (either **High** or **Bimodal** categories) in the species with higher CT~min~, *A. carolinensis*, relative to the null hypothesis of equal marginal frequencies. The dashed boxes highlight cells that would indicate matched responses in the two species, and the color of each cell (blue = excess, orange = deficit) represents the deviation of the observed from expected number of transcripts. The (A) *enhanced response* hypothesis proposes that the increase in cold tolerance in *A. picea* is achieved by amplifying existing molecular mechanisms, and thus there should be an excess of shared response types between species. In contrast, the (B) *tolerance hypothesis* predicts that *A. picea* is less reliant on induced responses to confer cold-tolerance than *A. carolinensis*, leading to an excess of shifts from induction in *A. carolinensis* to the **Not Responsive** or down-regulation categories in *A. picea*.

**Figure 2**. Results of marginal frequencies analysis of thermal response categories. The color of each cell  (blue = excess, orange = deficit) represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies. The units are number of transcripts. For each temperature extreme, the species expected to be less well adapated to that extreme is displayed on the x-axis for the two response categories corresponding to upregulation (**Bimodal** and **Low** for the low temperatures, or **Bimodal** and **High** for high temperatures). The distribution of response categories for those transcripts in the better-adapted species is arrayed along the y-axis. The dashed boxes indicate the matched responses (e.g. **High** - **High**). (A) Low temperature extreme: there is an excess of shared **Low** and **Bimodal** expression types and a bias away from all other categories in *A. picea*, consistent with the *enhanced response* hypothesis (Fig. 1). (B) High temperature extreme: in addition to an excess of matched categories, there is an excess of **High** and **Bimodal** transcripts in *A. picea* that are not upregulated in *A. carolinensis* (**Intermediate** and **Not Responsive**), partially consistent with the *tolerance* hypothesis. The complete set of matched observations is shown in Fig. S1. Expression types are defined in Table 1.

@@

# Tables and Figures

```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally-responsive transcripts by expression type for *A. carolinensis* and *A. picea*. **Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are not thermally responsive in one species but are in the other species."}

>>>>

**Background** The distributions of species and their responses to climate change are in part determined by their thermal tolerances. However, little is known about how thermal tolerance evolves. To test whether evolutionary extension of thermal limits is accomplished through enhanced cellular stress response (*enhanced response*), constitutively elevated expression of protective genes (*genetic assimilation*) or a shift from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*: the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized thermal reactionomes of two common ant species in the eastern U.S, the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*, across 12 temperatures that spanned their entire thermal breadth.

**Results** We found that at least 2% of all genes were thermally responsive. The majority of upregulation was specific to exposure to low temperatures. The cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than did *A. carolinensis*, consistent with the *enhanced response* hypothesis. In contrast, under high temperatures the warm-adapted *A. carolinensis* downregulated many of the genes upregulated in *A. picea*, and required more extreme temperatures to induce down-regulation in gene expression, consistent with the *tolerance* hypothesis. We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*.

**Conclusions** These results suggest that increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.

@@

The combined *Aphaenogaster* transcriptome assembly contained `r prettyNum(nrow(annotation.table), big.mark =",")` transcripts. About half of these (`r prettyNum(nrow(annotation.table.responsive), big.mark =",")`) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. We fit polynomial regression models to the expression values of each transcript to identify those that had a linear or quadratic relationship. To account for multiple tests, we both applied a False Discovery Rate (FDR) correction, and performed a resampling analysis to determine the number of transcripts that would be expected to have a significant relationshp by chance alone. We retained the `r nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)` (`r round((nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)) / nrow(annotation.table), 3) * 100`% of total) true positive transcripts that exceeded null expectation from the resampling analysis. Of these transcripts, `r round(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"] / sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), 2) *100`% (`r prettyNum(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"], big.mark=",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g. high vs. low temperature). The proportion of responsive transcripts is similar if we focus only on those transcript with a BLAST hit (`r prettyNum(responsive_transcripts_w_BLAST, big.mark = ",")` significant transcripts out of `r prettyNum(nrow(annotation.table.responsive), big.mark = ",")`). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] so this estimate is likely a lower bound for the true proportion of transcripts that are thermally responsive.

@@

In addition, we also examined the specific patterns of shifts from one expression category to another between species. When controlling for overall differences in the number of responsive transcripts, the expression categories of individual transcripts between the two species were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 319, P < 0.001, Fig. S1). Specifically, the two species differed significantly in expression pattern, which captures differences in slope as well as category, for `r prettyNum(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"], big.mark = ",")` (`r round(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"] / sigtable[which(sigtable$Coefficient=="Total"), "Number.significant"], 2) * 100`%) of the thermally responsive transcripts. Most of these shifted from one to another expression category.

@@

Similarly, for down-regulated (**Intermediate**) transcripts, we compared the mean temperatures of downregulation of transcript expression between species at both high (> 20°C) and low (< 20°C) temperatures. Consistent with the *tolerance hypothesis*, *A. carolinensis* had greater inertia of gene expression in response to increasing temperatures. The temperature of downregulation for **Intermediate** transcripts was `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`). The difference between species was not significant with decreasing temperatures (*t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig. 3B).

@@

We tested the *genetic assimilation hypothesis* by comparing the log ratios of relative inducibility to relative baseline expression at the rearing temperature (25°C). If stress-response transcripts have shifted between species from inducible to constitutive expression, there should be a negative relationship between the two. We found no evidence of such a relationship for either temperature extreme: transcripts more upregulated at high temperatures in the cool-climate *A. picea* were not expressed at higher baseline levels in the warm-climate *A. carolinensis* (Fig. 4A). Similarly, transcripts more upregulated at low temperatures in *A. carolinensis* did not show higher baseline levels in *A. picea* (Fig. 4B). In fact, for both comparisons we found a weakly positive relationship between relative inducibility and baseline expression between the two species ($\beta_1$ = 0.31, *P* < 0.001 and ($\beta_1$ = 0.21, *P* < 0.001). In addition, the thermally responsive transcripts in *A. carolinensis*, regardless of expression pattern, had higher baseline expression than those in *A. picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`). An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A. picea* but were not thermally responsive in *A. carolinensis* (top-row of Fig. 2B). These transcripts are less likely to be upregulated in any context, consistent with the *tolerance hypothesis*.

@@

To limit differences in gene expression not related to the experimental treatment, on 12 different days we haphazardly collected three ants from each 2012 colony at the same time of day to minimize variation due to circadian oscillations. We measured response to temperature with a one-hour static temperature application, which is ecologically relevant for workers that leave the thermally-buffered nest and are immediately exposed to ambient temperatures while foraging [@lubertazzi2012]. Each day, the ants were placed in glass tubes immersed in a water bath maintained at one of 12 randomly-assigned temperatures (0° to 38.5°C, in 3.5° increments) for one hour. The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum (~0°C) and maximum (~43°C) temperatures for *Aphaenogaster* [@warren2013], and these treatments did not cause mortality. At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C. Thus, our reactionome characterized early, but not late, responding genes. We extracted mRNA by homogenizing the three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.

@@

We assembled the sequenced reads into the full set of mRNA transcripts, the transcriptome, for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011]. *De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g. regions of two separate transcripts that assemble into a false, or chimeric, third transcript) [@yang2013]. We removed potentially false transcripts by first running the program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length. Second, we ran the program `uclust` which clusters sequences completely contained within longer sequences at greater than 90% similarity (see Supplementary Methods). We used liberal values (90% similarity) to merge orthologous transcripts in the two species that may not have assembled together in the initial *de novo* transcriptome assembly. To identify contaminant sequences, we screened our full transcriptome using the program `DeconSeq` [@schmieder2011] with the provided bacteria, virus, archaen and human databases of contaminants.

@@

We quantified expression of each transcript using the program `Sailfish` [@patro2014] and used the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of transcript expression. We included the contaminant transcripts identified by `DeconSeq` at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses. Because preliminary examination of the data (Supplementary Methods) indicated that the 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplementary Methods).

@@

All analyses were performed with R 3.2 [@r_core_team2013] and are fully reproducible (Supplementary Methods).

@@

Table S1 contains the results of the gene set enrichment analysis, showing the enriched gene ontology terms for each species in each thermal response category.

The Supplementary Methods contain the detailed information on the analysis. The reproducible and version-controlled scripts underlying the analysis are available on GitHub (https://github.com/johnstantongeddes/ApTranscriptome).

@@

## Competing interests

The authors declare they have no competing interests.

## Authors' contributions

@@

**Figure 1**. Illustrations of the expected thermal response patterns in the two species under alternative mechanistic hypotheses of temperature adaptation. Although both temperature extremes were investigated in a similar way, for simplicity only the response to low temperatures is illustrated here. Each column indicates the distribution across all response categories in *A. picea*, which has a lower CT~min~ and is therefore better adapted to low temperatures, for the set of transcripts identified as cold-induced (either **High** or **Bimodal** categories) in the species with higher CT~min~, *A. carolinensis*, relative to the null hypothesis of equal marginal frequencies. The dashed boxes highlight cells that would indicate matched responses in the two species, and the color of each cell (blue = excess, orange = deficit) represents the deviation of the observed from expected number of transcripts. The (A) *enhanced response* hypothesis proposes that the increase in cold tolerance in *A. picea* is achieved by amplifying existing molecular mechanisms, and thus there should be an excess of shared response types between species. In contrast, the (B) *tolerance hypothesis* predicts that *A. picea* is less reliant on induced responses to confer cold-tolerance than *A. carolinensis*, leading to an excess of shifts from induction in *A. carolinensis* to the **Not Responsive** or down-regulation categories in *A. picea*.

**Figure 2**. Results of analysis of thermal response patterns in the two species. The color of each cell  (blue = excess, orange = deficit) represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies. The units are number of transcripts. For each temperature extreme, the species expected to be less well adapated to that extreme is displayed on the x-axis for the two response categories corresponding to upregulation (**Bimodal** and **Low** for the low temperatures, or **Bimodal** and **High** for high temperatures). The distribution of response categories for those transcripts in the better-adapted species is arrayed along the y-axis. The dashed boxes indicate the matched responses (e.g. **High** - **High**). (A) Low temperature extreme: there is an excess of shared **Low** and **Bimodal** expression types and a bias away from all other categories in *A. picea*, consistent with the *enhanced response* hypothesis (Fig. 1). (B) High temperature extreme: in addition to an excess of matched categories, there is an excess of **High** and **Bimodal** transcripts in *A. picea* that are not upregulated in *A. carolinensis* (**Intermediate** and **Not Responsive**), partially consistent with the *tolerance* hypothesis. The complete set of matched observations is shown in Fig. S1. Expression types are defined in Table 1.

@@

# Tables

```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally responsive transcripts by expression type for *A. carolinensis* and *A. picea*. **Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are not thermally responsive in one species but are in the other species."}
